Design and Implementation of Probabilistic Programming Language Anglican

### 2016-08-18 paper/paper.tex
pre-submission draft

\newcommand*{\dt}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf DT:} 
        {\rm #1}\end{minipage}}}
\newcommand*{\fw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf FW:} 
        {\rm #1}\end{minipage}}}
\newcommand*{\hy}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:} 
        {\rm #1}\end{minipage}}}
\newcommand*{\jw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf JW:} 
        {\rm #1}\end{minipage}}}

@@

% \hy{Great introduction. It very nicely explains why Anglican
% is born and how it is implemented. If I act as a really picky
% reviewer or a reader, I would also expect the introduction
% to help me grasp in a high level what new insights or scientific problems 
% this development of Anglican have resulted in or solved so far.}

@@

Compilation of Anglican into Clojure relies on the Clojure
\textit{macro} facility. However, the compilation algorithm is
implemented as a library of \textit{functions} in namespace
\texttt{anglican.trap}, which are invoked by macros.
The CPS transformation is organized in top-down manner. The
top-level function is  \texttt{cps-of-expression}, which
receives an expression and a continuation, and returns the
expression in the CPS form, with the computed result passed to
the continuation. A continuation accepts \textit{two} arguments:
\begin{itemize}
    \item the computed value;
    \item the internal state, bound to the local variable
        \texttt{\$state} in every lexical scope.
\end{itemize}
The CPS transformation of constant \texttt{1} with continuation
\texttt{cont} thus takes the following form:
\begin{lstlisting}[style=default]
=> (cps-of-expression 1 'cont)
(cont 1 $state)
\end{lstlisting}

@@

@@

There are two proper probabilistic forms turning Anglican into a
probabilistic programming language --- \texttt{sample} and
\texttt{observe}. Their purpose is to interrupt deterministic
computation and transfer control to the inference algorithm.
Practically, this is achieved through returning
\textbf{checkpoints} --- Clojure records of the corresponding
types (\texttt{anglican.trap.sample} or
\texttt{anglican.trap.observe}). The records contain fields
specific to each form, as well as the continuation; calling the
continuation resumes the computation. Checkpoints expose the
program state to the inference algorithm, and the updated state
is re-injected into the computation when the continuation is
called: 
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist val) 'cont)
(->observe dist val cont $state)
\end{lstlisting}

In addition to checkpoints, there are a few other special forms
--- \texttt{predict}, \texttt{store}, \texttt{retrieve},
\texttt{mem} --- which modify program state. These forms are
translated into expressions involving calls of functions from the
\texttt{anglican.state} namespace. The \texttt{mem} form, which
implements stochastic memoization, deserves a more detailed
explanation.

\textbf{Memoization} is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program, hence \texttt{mem}'s
memory is stored as a nested dictionary in the program state
(function \texttt{mem-cps}).  Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn []
        (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply
        foo
        ;; memoize result in continuation
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state
        P)))))
\end{lstlisting}

@@

\hy{I rewrote various parts so as to avoid the word multimethod.
My strategy was to use the phrases, such as (ad-hoc) polymophic
function and overloading. All of my rephrasing should be
double-checked.}

@@

Macro \texttt{defdist} should be used to define distributions.
\texttt{defdist} takes care of defining a separate record type
for every distribution so that multimethods can be dispatched on
distribution types when needed, e.g. for custom proposals. The
Bernoulli distribution could be defined as follows:
\begin{lstlisting}[style=default]
(defdist bernoulli
  "Bernoulli distribution"
  [p]
  (sample* [this] (if (< (rand) p) 1 0))
  (observe* [this value]
    (Math/log (case value
                1 p
                0 (- 1. p)
                0.))))
\end{lstlisting}

>>>>

\newcommand*{\dt}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf DT:} 
        {\rm #1}\end{minipage}}}
\newcommand*{\fw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf FW:} 
        {\rm #1}\end{minipage}}}
\newcommand*{\hy}[1]{}
% {\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:} 
%        {\rm #1}\end{minipage}}}
\newcommand*{\jw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf JW:} 
        {\rm #1}\end{minipage}}}

@@

\hy{Great introduction. It very nicely explains why Anglican
is born and how it is implemented. If I act as a really picky
reviewer or a reader, I would also expect the introduction
to help me grasp in a high level what new insights or scientific problems 
this development of Anglican have resulted in or solved so far.}

@@

\cbstart
Compilation of Anglican into Clojure is built around a variant of CPS
transformation. In a basic CPS-transformed program, a
continuation receives a single argument --- the computed value.
In Anglican, there are two flows of computation going in
parallel: values are computed by functional code, and,
at the same time, the state of the probabilistic program,
used by inference algorithms, is updated by 
probabilistic forms. Because of that, in Anglican a continuation
accepts  \textit{two} arguments:
\begin{itemize}
    \item the computed value;
    \item the internal state, bound to the local variable
        \texttt{\$state} in every lexical scope.
\end{itemize}

The compilation relies on the Clojure \textit{macro} facility,
and implemented as a library of \textit{functions} in namespace
\texttt{anglican.trap}, which are invoked by macros.  The CPS
transformation is organized in top-down manner.  The top-level
function is  \texttt{cps-of-expression}, which receives an
expression and a continuation, and returns the expression in the
CPS form.  For example, the CPS transformation of constant
\texttt{1} with continuation \texttt{cont} thus takes the
following form:
\begin{lstlisting}[style=default]
=> (cps-of-expression 1 'cont)
(cont 1 $state)
\end{lstlisting}

\subsection {The State}
\cbend

@@

\cbstart
So far we described the basics of Anglican compilation to
Clojure. The described approaches and techniques are important
for grasping the language implementation but relatively well-known.
In the rest of the section we focus on challenges we met and
resolved while implementing Anglican, as well as on
implementation of unique features of Anglican as a probabilistic
programming language.
\cbend

@@

There are two proper probabilistic forms turning Anglican into a
probabilistic programming language --- \texttt{sample} and
\texttt{observe}. Their purpose is to interrupt deterministic
computation and transfer control to the inference algorithm.
Practically, this is achieved through returning
\cbstart
\textbf{checkpoints} --- Clojure values of the corresponding
types (\texttt{anglican.trap.sample} or
\texttt{anglican.trap.observe}). The types contain fields
\cbend
specific to each form, as well as the continuation; calling the
continuation resumes the computation. Checkpoints expose the
program state to the inference algorithm, and the updated state
is re-injected into the computation when the continuation is
called: 
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist val) 'cont)
(->observe dist val cont $state)
\end{lstlisting}

In addition to checkpoints, there are a few other special forms
--- \texttt{predict}, \texttt{store}, \texttt{retrieve},
\texttt{mem} --- which modify program state. These forms are
translated into expressions involving calls of functions from the
\texttt{anglican.state} namespace. The \texttt{mem} form, which
implements memoization, deserves a more detailed explanation.

\cbstart
\paragraph{Memoization}
The author of a probabilistic model might want to draw a
feature of an entity randomly, but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved 
memoized functions. In Anglican, one might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person] 
                       (sample 
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (predict (eye-color 'bill))
    (predict (eye-color 'john))))
\end{lstlisting}
\cbend
Memoization is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program, hence \texttt{mem}'s
memory is stored as a nested dictionary in the program state
(function \texttt{mem-cps}).  Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn []
        (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply
        foo
        ;; memoize result in continuation
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state
        P)))))
\end{lstlisting}

@@

@@

Macro \texttt{defdist} should be used to define distributions.
\cbstart
\texttt{defdist} takes care of defining a separate type
\cbend
for every distribution so that multimethods can be dispatched on
distribution types when needed, e.g. for custom proposals. The
Bernoulli distribution could be defined as follows:
\begin{lstlisting}[style=default]
(defdist bernoulli
  "Bernoulli distribution"
  [p]
  (sample* [this] (if (< (rand) p) 1 0))
  (observe* [this value]
    (Math/log (case value
                1 p
                0 (- 1. p)
                0.))))
\end{lstlisting}
